{memes}: an R interface to the MEME Suite

### 2020-07-28 manuscript.Rmd
updating design & implementation

@@

The key underlying design principle of `memes` was to create tools that easily
integrate into the existing suite of tools available in `Bioconductor` and the
broader `R` package landscape. In this way, `memes` tools can be used to extend
existing analysis pipelines, or to create novel analyses.

>>>>

Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers which can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, thus the ability to
identify, classify, and compare motifs represents a key step in the analysis of
biological sequences. Although tools that implement these features have
historically existed in the Bioconductor landscape, they often rely on
proprietary data types with poor interoperability with other tools. Furthermore,
several of these packages will be deprecated in the next Bioconductor release,
leaving a gap in functionality with no clear alternatives.

The MEME Suite is a series of commandline and web-based tools which provide a variety
of features for motif analysis such as *de-novo* motif discovery, enrichment
testing, comparison, and matching \cite{meme-suite}. The MEME Suite is a widely
accepted set of tools with over 22,000 citations to date, and over 30,000 unique
users of the webserver implementation annually \cite{meme-usage-report-page,
google-scholar-meme}.

The MEME Suite family of tools provides a
comprehensive set of utilities to interrogate motif content across a broad range
of biological contexts. However, several factors limit the full potential of
these tools for use in data analysis. 
  1) must do significant preprocessing to get data in correct input format
  2) Easily parseable output data often doesn't contain all information
    - ex tomtom alternative hits in .tsv output only present in IUPAC format, not matrix
      - matrix is present in XML file, but no tools exist to parse this
  3) limited data visualization capabilities

Here we present `memes`, a motif analysis package which provides a seamless
interface to MEME Suite tools within R. `memes` uses base R and Bioconductor
data types for data input and output, facilitating better integration with
common analysis tools like the `tidyverse` and other Bioconductor packages.
Unlike the commandline implementation, `memes` outputs also function as inputs
to other MEME Suite tools, allowing simple construction of motif analysis
pipelines without additional data processing steps. Additionally, R/Bioconductor data
structures provide a full-featured representaiton of MEME Suite output data, providing
users quick access to all relevant data structures with simple syntax.

In addition to wrapping MEME Suite utilities, `memes` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow motif databases
to integrate external data to aid analysis, such as filtering a set of
transcription factor motifs based on gene expression in the tissue of interest.

`memes` also provides simple utilities for data visualization where appropriate.
Providing R representations of MEME Suite data objects enables generation of
novel data visualizations using the well established tools available in the R
landscape. Therefore, `memes` empowers users to deeply analyze and visualize
motif analysis data.

@@

### Datatypes

MEME Suite tools are run on the commandline and use files stored on-disk as
input while returning a series of output files containing varying data types.
`memes` reduces each MEME Suite tool to an R function which internally handles
file I/O such that all MEME Suite calling functions accept R objects as input,
and return R objects as output. Sequence inputs are stored in `BioStrings`
format, an R/Bioconductor package for storing biopolymer sequence data. Motif
inputs are passed as `universalmotif` objects, another R/Bioconductor package
for representing motif matrices and their associated metadata. `memes` will also
always accept paths to files on disk, such as fasta files for sequence inputs,
and meme format files for motif inputs.

The MEME Suite tools do not return consistent data structures across tools. For
example, the tool TomTom, which compares query motifs to a database of known
motifs, returns tab-separated results which do not contain the frequency matrix
representations of the matched motifs. Instead, users must parse these matrices
back out from the input databases, creating additional friction for creating
novel data visualizations and exploratory analyses. Often, these data are
contained in XML formatted outputs, which although they hold all relevant
information, are difficult to parse. To our knowledge, no publicly available
tools exist to parse MEME Suite XML formatted data, therefore users must write
custom code to extract these data. `memes` exports data import functions for
each supported MEME Suite tool, which import these data as R data.frames. These
functions work internally to create the seamless file I/O when calling MEME
Suite tools, but the import functions can be called directly by users to import
previously generated data for use in R.

Many of the MEME Suite tools, such as MEME, DREME, or TomTom return data which
describe properties of motifs. For example, MEME and DREME return de-novo
discovered motifs from query regions along with statistical information about
their enrichment. The `universalmotif` data structure is a powerful
R/Bioconductor representation for motif matrices and their associated metadata. 
However, `universalmotif` objects are not ideal for performing bulk manipulation
or queries of metadata, nor do they support complex data structures such as
nested elements. Many MEME Suite analysis pipelines involve identifying motifs
which are further analyzed by multiple tools, creating additional metadata. Thus
an ideal representation for these data is one which can hold an unlimited amount
of metadata, and contain complex heirarchies. `memes` therefore extends the
`universalmotif` format by providing data.frames holding motif metadata which
retain links to their `universalmotif` representation. These universalmotif
data.frames combine the advantages of the `universalmotif` framework for motif
manipulation with the extensibility and flexibility of R data.frames, and are
the basis for a majority of `memes` data outputs. These structures are also
valid input types to `memes` functions, and when used as such, output data are
appended as new columns to the input data, ensuring data provenance.

**Figure?: showing how the data.frame connects to universalmotif**

`universalmotif` data.frames can additionally be used to examine or manipulate
the properties of any motif database. Using `universalmotif` functions to import
motif data, `as_universalmotif_dataframe()` converts any `universalmotif` object
into a `universalmotif` data.frame which can be manipulated using base R or
`tidyverse` functions, just like regular R data.frames. Manipulation of linked
metadata columns will result in updating the linked `universalmotif` object when
calling `update_motifs()` to retain the data.frame structure, or when calling
`as_universalmotif()` to convert the data.frame into `universalmotif` format. In
this way, `memes` enables large-scale motif database manipulation & analysis
within the preexisting R framework.

### Genomic range-based data as entrypoint

### Old Design & Implementation
The key underlying design principle of `memes` was to create tools that easily
integrate into the existing suite of tools available in `Bioconductor` and the
broader `R` package landscape. In this way, `memes` tools can be used to extend
existing analysis pipelines, or to create novel analyses.
